---
title: "Multivariate Medical Outcome Analysis"
author: "Michael J Kesling, Ph.D."
output: rmarkdown::github_document
---
# Overview
This project aims to determine whether a particular medical treatment improves health, as measured by (1) weight loss, and (2) self-reported health (SRH) improvement.  All starting data have already been anonymized.  

This webpage omits a few lines of code in order to reduce error messages, etc.  The full R Markdown file can be seen [here].

```{r setup, include=FALSE, fig.height=8, fig.width=8, dpi=300}
knitr::opts_chunk$set(echo = TRUE)
require(dplyr)
require(magrittr)
require(ggplot2)
```

## Part 1: Creating an Analytical Dataset
The given data reside in 4 separate files.  To analyze the data, I'll create a single dataframe with one subject per line and one column per measurement.  

I start by reading in each file separately as a dataframe, cleaning it up, and then merging the 4 dataframes into a single one.  
```{r}
palette <- c("#000000", "#56B4E9")

# read in first file:
one <- read.csv("data1_age_sex_race.csv")

# separate sex, age and race data:
sex <- gsub("([A-Z]+)[0-9]+.*","\\1",one$SexAge_Race)
age <- gsub("[A-Z]+([0-9.]+).*","\\1",one$SexAge_Race)
race <- gsub("[A-Z]+[0-9.]+_(.*)","\\1",one$SexAge_Race)

# create main dataframe:
DF <- data.frame(cbind(subjID=one$subject_ID, sex, age, race))
```
Next, I process the second file.  *SRH* means Self-Reported Health.
```{r}
# read in second file
two <- read.csv("data2_SRH.csv")

# create pre-SRH and post-SRH columns and add to dataframe:
trt_preSRH <- two %>% filter(TIME == "PRE") %>% select(id, trt, SRH) %>%
  rename(preSRH = SRH)
postSRH <- two %>% filter(TIME == "POST") %>% select(id, SRH) %>%
  rename(postSRH = SRH)
```
I noticed that there were >1 entries for some patrients, and I needed 
to see if the data were simply duplicated or if there were inconsistencies:
```{r}
dupsPRE <- which(trt_preSRH$id %in% names(sort(table(trt_preSRH$id), decreasing =TRUE))[1:18])
trt_preSRH[dupsPRE,]
```
```{r}
dupsPOST <- which(postSRH$id %in% names(sort(table(postSRH$id), decreasing =TRUE))[1:17])
postSRH[dupsPOST,]
```
In all cases, the multiple entries for each patient are redundant. I can therefore randomly pick a single example from each patient.
```{r}
selectSingleExample <- function(df){
  alreadySeen <- df$id[1]
  cleaned <- df[1,]
  for(i in 2:dim(df)[1]){
    # code less general but faster b/c duplicates always consecutive:
    if(df$id[i] != alreadySeen[length(alreadySeen)]){
      alreadySeen <- c(alreadySeen, df$id[i])
      cleaned <- rbind(cleaned, df[i,])
    }
  }
  return(cleaned)
}

trt_preSRH_cleaned <- selectSingleExample(trt_preSRH)
postSRH_cleaned <- selectSingleExample(postSRH)

# check that order to patients has not changed between 3 matrices:
all(trt_preSRH_cleaned$id == postSRH_cleaned$id & DF$subjID == postSRH_cleaned$id)
```
The order of all patients was maintained during the duplicate-removal process.
```{r}
# remove redundant subject ID columns:
trt_preSRH_cleaned <- trt_preSRH_cleaned %>% select(trt, preSRH)
postSRH_cleaned <- postSRH_cleaned %>% select(postSRH)
DF <- cbind(DF, trt_preSRH_cleaned, postSRH_cleaned)
```

Next, I'll import files 3A and 3B, process them, and add the data to DF

```{r}
threeA <- read.csv("data3a_weight_trt.csv")
threeB <- read.csv("data3b_weight_con.csv")
colnames(threeB)[1] <- "Id"
three <- rbind(threeA, threeB)
# (all weight data now in a single data frame)

preWeight <- three %>% select(Id, PRE_WEIGHT) %>%
  filter(!is.na(PRE_WEIGHT)) %>% rename(id=Id)
postWeight <- three %>% select(Id, POST_WEIGHT) %>%
  filter(!is.na(POST_WEIGHT)) %>% rename(id=Id)


# Let's establish if the multiple-entries per patient are duplicates or not
preWeight[which(preWeight$id %in% names(sort(table(preWeight$id), decreasing =TRUE))[1:14]),]
```
```{r}
postWeight[which(postWeight$id %in% names(sort(table(postWeight$id), decreasing =TRUE))[1:15]),]
```
We can see that in the case of the pre- and post-weights, the multiple entries are not precisely duplicate entries.  However, the multiple-values per patient are all within about 0.1% of each other.  Therefore, I'll simply select the first entry in each case:
```{r}
preWeight_cleaned <- selectSingleExample(preWeight)
postWeight_cleaned <- selectSingleExample(postWeight)

# next, I'll test that the order of subject ID's has not changed:
all(preWeight_cleaned$id == postWeight_cleaned$id & postWeight_cleaned$id == DF$subjID)
```
We see that the order of all subjects has been maintained.  
We next add the additional weight data to the dataframe DF:
```{r}
# remove un-needed "id" columns:
preWeight_cleaned <- preWeight_cleaned %>% select(PRE_WEIGHT)
postWeight_cleaned <- postWeight_cleaned %>% select(POST_WEIGHT)

# finish merging the rest of the data into DF
DF <- cbind(DF, preWeight_cleaned, postWeight_cleaned)
```
Lastly, we convert *age* and *subject* to numeric variables  
```{r}
DF$age <- as.numeric(as.character(DF$age))
DF$subjID <- as.numeric(as.character(DF$subjID))
lapply(DF, class)
```
### Converting Weight and SRH Variables to Difference Variables
This study aims to see if the treatment has an effect on either altering the patients weight or on altering the Self Reported Health.  So I'm going to subtract the initial weight from the final weight. I'm going to convert the SRH vars to numeric vars:  
1. Very Poor  
2. Poor  
3. Satisfactory  
4. Good  
5. Excellent  
and perform a post-SRH minus pre-SRH subtraction as well.  It should be noted that the "Excellent" and "Very Poor" levels appear more than once due to space characters.  These will be cleaned up as well.
```{r}
# clean up typographical errors
DF$preSRH <- gsub("Excellent ", "Excellent", DF$preSRH)
DF$preSRH <- gsub("Very  Poor", "Very Poor", DF$preSRH)
DF$postSRH <- gsub("Excellent ", "Excellent", DF$postSRH)
DF$postSRH <- gsub("Very  Poor", "Very Poor", DF$postSRH)
unique(DF$preSRH)
unique(DF$postSRH)


# Subtract pre- from post-SRH
DF$preSRH <- as.numeric(DF$preSRH %>% {gsub("Excellent", 5, .)} %>% 
  {gsub("Good", 4, .)} %>% {gsub("Satisfactory", 3, .)} %>% 
  {gsub("Very Poor", 1, .)} %>% {gsub("Poor", 2, .)})

DF$postSRH <- as.numeric(DF$postSRH %>% {gsub("Excellent", 5, .)} %>% 
  {gsub("Good", 4, .)} %>% {gsub("Satisfactory", 3, .)} %>% 
  {gsub("Very Poor", 1, .)} %>% {gsub("Poor", 2, .)})


# add new columns to the dataframe DF
DF <- DF %>% mutate(weightGain=POST_WEIGHT - PRE_WEIGHT)
DF <- DF %>% mutate(SRHimprove=postSRH - preSRH)
```
## Part 2: Analyzing Study from Weight Loss Vantage Point

### Plot of Weight Loss by Treatment
We start off by simply plotting *weightGain* vs *treatment*  
```{r, dpi=300}
ggplot(DF, aes(x=trt, y=weightGain)) +
  geom_jitter(size=0.5) + xlab("Treatment") + ylab("Weight Change")
```
We see 2 trends immediately.  The control subjects had remarkably little change in their weights, but the treated subjects had (1) much greater variance in their weight change, and (2) a net loss in weight.  It's also clear that in some cases, the treatment caused weight gain.

### Simple Linear Regression of Weight Change
We perform a simple linear regression on treatment on weight gain to observe the magnitude and statistical significance of the treatment.  
```{R}
weightModel <- lm(weightGain ~ trt, data=DF)
summary(weightModel)
```
We see that, on average, there's an 8.2 pound weight loss amongst the treated subjects compared to the control subjects, with a small standard error (0.21).  The t-statisic shows that we're over 45 standard errors away from the null model of no change, and the p-value is well under 2e-16.  

The adjusted R-squared value says that the treatment variable only explains 29% of the variance.  Nevertheless, we have plenty of statistical power supporting the treatment variable acting as a predictor for *weightGain* in this simple linear regression.

### Multivariate Modeling of Weight Loss  
We'll now build multivariate models using *sex*, *race*, and *treatment* to see how these variables affect weight loss.  We'd also like to see if we can understand anything about those who are actually gaining weight as the result of the treatment.  
```{r}
weightMVR <- lm(weightGain ~ sex * race * trt, data=DF)
summary(weightMVR)
```
We see that *overall* men lose about 2 pounds more than women, when their treatment/control status is not included.  We see that those in the treatment group lose almost 3 pounds compared to the control group, when no other attributes are included.  
We also see a specific Male:Treatment interaction term that is quite large.  Men who are treated lose an additional 10 pounds above and beyond the 2 pound loss of being men and the 3 pound loss of being treated.  
All 3 of these coefficients are statistically significant.  
The variables and all the interaction terms now explain about 65% of the weightGain variance.  
Neither *age* (not shown) nor *race* have a statistically significant coefficient.  The age-sexMALE interaction term has a p-value of about 0.06, but the magnitude is only -0.06 pounds.  

### Analyzing Weight Loss in Men and Women Separately
I'm going to follow this up with a plot, separating out men and women.  
```{r, dpi=300}
ggplot(DF, aes(x=sex, y=weightGain), colour=trt) +
  geom_jitter(aes(colour=as.factor(trt)), size=0.5) + 
  ggtitle("Weight Change for Treated and Control Women and Men") +
  ylab("Weight Gain") +
  scale_color_manual(name="Treatment",
                      breaks = c("0","1"),
                      values = palette,
                      labels = c("No", "Yes"))
```
This plot is actually quite remarkable, as it shows that virtually all weight gain that happens, happens to treated women.  It also shows that men respond much better in terms of weight loss--both the percentage of men that lose weight and the magnitude of the weight loss.

### Distribution of Weight Change by Sex and Treatment
I'm going to plot the densities of the weight change broken down by sex and treatment, which is simply another representation of what the previous figure showed.  
```{r, dpi=300}
labels <-c("0"="Control Subjects", "1"="Treated Subjects")
ggplot(DF, aes(x=weightGain)) + 
   geom_line(aes(colour=sex), stat="density") +
   facet_wrap(~ trt, ncol=2, labeller = labeller(trt=labels)) +
   ggtitle("Weight Gain Density by Sex and Treatment")

```
The *weightGain* variable gives normal distributions when only control subjects are considered and when only treated subjects are considered.  So we can use the *weightGain* variable as-is.  We can also see clearly that the variance for both sexes is small and essentially equal for the control subjects and is large and essentially equal for treatment subjects.  The mean for men in both groups is lower than for women.  


### Modeling Men 
If we take sex as a known and build separate models for women and for men, I suspect that the simple linear regression model for men will be extremely predictive.

```{r}
DF_men <- DF %>% filter(sex == "MALE")
modelMen <- lm(weightGain ~ trt, data=DF_men)
summary(modelMen)
```
The treatment is associated with a 13.6 pound weight loss in men, and it's over 70 standard errors from a weight loss of zero.  *treatment* also explains nearly 70% of the variance for men.  

### Modeling Women
```{r}
DF_women <- DF %>% filter(sex == "FEMALE")
modelWomen <- lm(weightGain ~ trt, data=DF_women)
summary(modelWomen)
```
When the model looks only at women, there is an average weight loss of about 3 pounds, and it's quite statistically significant.  However, *treatment* only explains about 9% of the variance for women.

### Looking at Predictors More Carefully for Women
I'm going to further tease apart the Women dataset to see if I can isolate those who are gaining weight.  
```{r}
modelWomenMVR <- lm(weightGain ~ trt * age * race, data=DF_women)
summary(modelWomenMVR)
```
Neither *age* nor *race*, with or without *treatment* has any predictive power, and does not help separate women who gain weight from women who lose weight.  Creating smaller models with less predictors did not shed more light on this issue (not shown).  

It should be noted that each coefficient represents how much the weightGain variable changes per single unit change in the predictor, while holding all other predictors constant.  However, the predictors can be correlated with one another, and it may not be possible to only change the predictor in question while not changing any other predictors.  
Next, I'd like to plot out the women by race for a visual effect of what we just saw in the MVR model:  
```{R, dpi=300}
ggplot(DF_women, aes(x=race, y=weightGain), colour=trt) +
  geom_jitter(aes(colour=as.factor(trt)), size=0.5) +
  ggtitle("Weight Gain for Women by Race") +
  scale_color_manual(name="Treatment",
                      breaks = c("0","1"),
                      values = palette,
                      labels = c("No", "Yes"))
```

Plotting women's age:
```{r, dpi=300}
ggplot(DF_women, aes(x=age, y=weightGain), colour=trt) +
  geom_jitter(aes(colour=as.factor(trt)), size=0.5) +
  ggtitle("Weight Gain of Women Plotted vs. Age") +
  scale_color_manual(name="Treatment",
                      breaks = c("0","1"),
                      values = palette,
                      labels = c("No", "Yes"))
  
```
Again, consistent with the *modelWomenMVR* above, age is not an explanatory variable, with the possible exception of women under 25 and women over 55, had we broken *age* into 3 distinct ranges.  Overall, race and age have little effect on weight loss or gain for women.  


## Part 3: Analyzing Study from SRH Improvement Vantage Point

### SRH Simple Linear Regression
In terms of SRH, there does seem to be more density in the positive direction and less density in the negative direction for treated subjects compared to control subjects.

```{r}
SRHmodel <- lm(SRHimprove ~ trt, data=DF)
summary(SRHmodel)
```
We see that the treatment variable explains only about 4.5% of the total SRH improvement variance.  The average SRH improvement is 1/2 of a category.  We also have very strong statistical support.
```{r}
SRH_MVR <- lm(SRHimprove ~ trt * sex, data=DF)
summary(SRH_MVR)
```
Here, we're seeing that the treatment improves SRH overall, and there's an additional effect for treated men.  Each coefficient represents about a 0.3 category change (e.g. 30% the size from Satisfactory to Good).

```{r, dpi=300}
ggplot(DF, aes(x=trt, y=SRHimprove), colour=sex) + 
  geom_jitter(aes(colour=sex), size=0.5) +
  theme_bw() + ggtitle("Self Reported Health Improvement") +
  ylab("SRH improvement") + xlab("Treatment") +
  scale_color_manual(name="Sex",
                      breaks = c("MALE", "FEMALE"),
                      values = c("red", "blue"),
                      labels = c("Male", "Female"))
```
Once can clearly see the higher density of treated men compared to treated women with a positive SRH.  This is particularly apparent when the SRH improvement is 3 categories or greater.  

```{r}
srhSex <- lm(SRHimprove ~ trt * age, data=DF)
summary(srhSex)
```
We see that there's a significant age:treatment interaction term.
```{r, dpi=300}
ggplot(DF, aes(x=age, y=SRHimprove), colour=trt) +
  geom_jitter(aes(colour=as.factor(trt)), size=0.3) +
  ggtitle("SRH Improvement of Subjects vs. Age") +
  theme_bw() + ylab("SRH Improvement") +
  scale_color_manual(name="Treatment",
                      breaks = c("0","1"),
                      values = palette,
                      labels = c("no", "yes"))
  
```
Amongst the people whose SRH changes during the study,  
1. No one is under the age of 32
2. There is an enrichment of untreated people who get worse between ages 32-39.
3. There is an enrichment of treated people who get better between 45-50.
4. There is an enrichment of untreated people who get a little better between 50-55.
5. Overall, increasing age predicts SRH improvement for treated subjects.


Increasing age improves SRH improvement, but with a small magnitude.
Being hispanic or white also explains part of the improvement with larger magnitude but less statistical significance.
I'm going to now build separate SRH models for men and women.
```{r}
SRHmen <- lm(SRHimprove ~ trt, data=DF_men)
summary(SRHmen)
```
```{r}
SRHwomen <- lm(SRHimprove ~ trt, data=DF_women)
summary(SRHwomen)
```


### Conclusions (all statistically significant):
1. The treatment is very effective for weight loss in men, with an average loss of over 13 pounds.  
2. The treatment gives rise to an average weight loss of 3 pounds for women.  The treatment sometimes gives rise to weight gain for some women.  
3. Neither race nor age is a significant predictor for weight loss.
4. The treatment gave rise to a 1/3-unit improvement in self-reported health index in women and 2/3-unit improvement in men.  
5. Those below age 32 did not undergo any SRH changes--either positive or negative.
6. Increasing Age had a positive effect on SRH improvement for treated individuals.
