---
title: "Multivariate Medical Outcome Analysis"
author: "Michael J Kesling, Ph.D."
output: rmarkdown::github_document
---
# Overview
This project aims to determine whether a particular medical regimen improves health, as measured by weight loss and self-reported health (SRH).  All data have been anonymized.  

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Part 1: Creating an Analytical Dataset
The given data reside in 4 separate files.  To analyze the data, I'll create a single dataframe with one subject per line and one column per measurement.  

I start by reading in each file separately as a dataframe, cleaning it up, and then merging the 4 dataframes into a single one.  
```{r}
require(dplyr)
require(magrittr)
require(caTools)
require(ggplot2)
palette <- c("#000000", "#56B4E9")

# read in first file:
one <- read.csv("data1_age_sex_race.csv")

# separate sex, age and race data:
sex <- gsub("([A-Z]+)[0-9]+.*","\\1",one$SexAge_Race)
age <- gsub("[A-Z]+([0-9.]+).*","\\1",one$SexAge_Race)
race <- gsub("[A-Z]+[0-9.]+_(.*)","\\1",one$SexAge_Race)

# create main dataframe:                 # MATRIX RIGHT NOW
DF <- data.frame(cbind(subjID=one$subject_ID, sex, age, race))
```
Next, I process the second file:
```{r}
# read in second file
two <- read.csv("data2_SRH.csv")

# create pre-SRH and post-SRH columns and add to dataframe:
trt_preSRH <- two %>% filter(TIME == "PRE") %>% select(id, trt, SRH) %>%
  rename(preSRH = SRH)
postSRH <- two %>% filter(TIME == "POST") %>% select(id, SRH) %>%
  rename(postSRH = SRH)
```
I noticed that there were >1 entries for some patrients, and I needed 
to see if the data were simply duplicated or if there were inconsistencies:
```{r}
dupsPRE <- which(trt_preSRH$id %in% names(sort(table(trt_preSRH$id), decreasing =TRUE))[1:18])
trt_preSRH[dupsPRE,]
```
```{r}
dupsPOST <- which(postSRH$id %in% names(sort(table(postSRH$id), decreasing =TRUE))[1:17])
postSRH[dupsPOST,]
```
In all cases, the multiple entries for each patient are redundant. I can therefore randomly pick a single example from each patient.
```{r}
selectSingleExample <- function(df){
  alreadySeen <- df$id[1]
  cleaned <- df[1,]
  for(i in 2:dim(df)[1]){
    # code less general but faster b/c duplicates always consecutive:
    if(df$id[i] != alreadySeen[length(alreadySeen)]){
      alreadySeen <- c(alreadySeen, df$id[i])
      cleaned <- rbind(cleaned, df[i,])
    }
  }
  return(cleaned)
}

trt_preSRH_cleaned <- selectSingleExample(trt_preSRH)
postSRH_cleaned <- selectSingleExample(postSRH)

# check that order to patients has not changed between 3 matrices:
all(trt_preSRH_cleaned$id == postSRH_cleaned$id & DF$subjID == postSRH_cleaned$id)
```
The order of all patients was maintained during the duplicate-removal process.
```{r}
# remove redundant subject ID columns:
trt_preSRH_cleaned <- trt_preSRH_cleaned %>% select(trt, preSRH)
postSRH_cleaned <- postSRH_cleaned %>% select(postSRH)
DF <- cbind(DF, trt_preSRH_cleaned, postSRH_cleaned)
```

Next, I'll import files 3A and 3B, process them, and add the data to DF

```{r}
threeA <- read.csv("data3a_weight_trt.csv")
threeB <- read.csv("data3b_weight_con.csv")
colnames(threeB)[1] <- "Id"
three <- rbind(threeA, threeB)
# (all weight data now in a single data frame)

preWeight <- three %>% select(Id, PRE_WEIGHT) %>%
  filter(!is.na(PRE_WEIGHT)) %>% rename(id=Id)
postWeight <- three %>% select(Id, POST_WEIGHT) %>%
  filter(!is.na(POST_WEIGHT)) %>% rename(id=Id)


# Let's establish if the multiple-entries per patient are duplicates or not
preWeight[which(preWeight$id %in% names(sort(table(preWeight$id), decreasing =TRUE))[1:14]),]
```
```{r}
postWeight[which(postWeight$id %in% names(sort(table(postWeight$id), decreasing =TRUE))[1:15]),]
```
We can see that in the case of the pre- and post-weights, the multiple entries are not precisely duplicate entries.  However, the multiple-values per patient are all within about 0.1% of each other.  Therefore, I'll simply select the first entry in each case:
```{r}
preWeight_cleaned <- selectSingleExample(preWeight)
postWeight_cleaned <- selectSingleExample(postWeight)

# next, I'll test that the order of subject ID's has not changed:
all(preWeight_cleaned$id == postWeight_cleaned$id & postWeight_cleaned$id == DF$subjID)
```
We see that the order of all subjects has been maintained.  
We next add the additional weight data to the dataframe DF:
```{r}
# remove un-needed "id" columns:
preWeight_cleaned <- preWeight_cleaned %>% select(PRE_WEIGHT)
postWeight_cleaned <- postWeight_cleaned %>% select(POST_WEIGHT)

# finish merging the rest of the data into DF
DF <- cbind(DF, preWeight_cleaned, postWeight_cleaned)
```
Lastly, age and subject are not stored as numeric
```{r}
DF$age <- as.numeric(as.character(DF$age))
DF$subjID <- as.numeric(as.character(DF$subjID))
lapply(DF, class)
```
### Converting Weight and SRH Variables to Difference Variables
This study aims to see if the treatment has an effect on either altering the patients weight or on altering the Self Reported Health.  So I'm going to subtract the initial weight from the final weight. I'm going to convert the SRH vars to numeric vars:  
1. Very Poor  
2. Poor  
3. Satisfactory  
4. Good  
5. Excellent  
and perform a post minus pre subtraction as well.  It should be noted that the "Excellent" and "Very Poor" levels appear more than once due to space characters.  These will be cleaned up as well.
```{r}
# clean up typographical errors
DF$preSRH <- gsub("Excellent ", "Excellent", DF$preSRH)
DF$preSRH <- gsub("Very  Poor", "Very Poor", DF$preSRH)
DF$postSRH <- gsub("Excellent ", "Excellent", DF$postSRH)
DF$postSRH <- gsub("Very  Poor", "Very Poor", DF$postSRH)
unique(DF$preSRH)
unique(DF$postSRH)

# barplot(prop.table(table(DF$preSRH)))
# barplot(prop.table(table(DF$postSRH)))

# I am going to assume, for the moment, that these 5 categories are equally spaced.
DF$preSRH <- as.numeric(DF$preSRH %>% {gsub("Excellent", 5, .)} %>% 
  {gsub("Good", 4, .)} %>% {gsub("Satisfactory", 3, .)} %>% 
  {gsub("Very Poor", 1, .)} %>% {gsub("Poor", 2, .)})

DF$postSRH <- as.numeric(DF$postSRH %>% {gsub("Excellent", 5, .)} %>% 
  {gsub("Good", 4, .)} %>% {gsub("Satisfactory", 3, .)} %>% 
  {gsub("Very Poor", 1, .)} %>% {gsub("Poor", 2, .)})


# now create new DF with 
DF <- DF %>% mutate(weightGain=POST_WEIGHT - PRE_WEIGHT)
DF <- DF %>% mutate(SRHimprove=postSRH - preSRH)
```
### Distribution of Output Variables
Ideally our difference-SRH and difference-weight variables will be approximately normally distributed.  Let's plot a histogram of each, using only the control subjects (*trt* = 0)  
```{r}
DFctrl <- DF %>% filter(trt == 0)
DFtrmt <- DF %>% filter(trt == 1)
par(c(2,1))
hist(DFctrl$weightGain)
hist(DFtrmt$weightGain)
#                             IMPROVE WITH GGPLOT BOTH AT ONCE
```
The *weightGain* variable gives normal distributions when only control subjects are considered and when only treated subjects are considered.  So we can use these *weightGain* variable as-is.  


### Dividing Data into Training and Test Sets
The models I'll build will be optimized for test data (using cross validation), rather than training data.  Now that our data have been cleaned, we'll randomly select 75% of the data as the training set and 25% of the data as the test set, each evenly split between no-treatment and treatment.  
```{r}
set.seed(38923)
idxTrain <- sample.split(DF$trt, SplitRatio = 0.75)
Train <- DF[idxTrain,]
Test <- DF[!idxTrain,]

Train_ctrl <- Train %>% filter(trt == 0)
Train_trmt <- Train %>% filter(trt == 1)
```

### Initial ANOVA Analysis
The simplest way to consider the effect of the treatment is to ignore *sex* and *race* and simply see what the 2-tailed 95% confidence intervals look like for *weightGain* vs treatment and *SRHimprove* vs treatment.

I'd like to start with 2 plots:
```{r}
require(ggplot2)
ggplot(Train, aes(x=trt, y=weightGain)) +
  geom_jitter(size=0.5)
```
We see 2 trends immediately.  The control subjects had remarkably little change in their weights, but the treated sample had much greater variance in their weight change, with the average being a weight loss.  It's also clear that in some cases, the treatment caused weight gain.
```{r}
ggplot(Train, aes(x=trt, y=SRHimprove), colour=sex) +
  geom_jitter(aes(colour=sex), size=0.5)
```
In terms of SRH, there does seem to be more density in the positive direction and less density in the negative direction for treated subjects compared to control subjects.

### Performing Linear Regression
```{R}
weightModel <- lm(Train$weightGain ~ Train$trt)
summary(weightModel)
```
We see that, on average, there's an 8.2 weight loss amongst the treated subjects compared to the control subjects, with a small standard error (0.21).  The t-statisic shows that we're about 40 standard errors away from the null model of no change, and the p-value is < 2e-16.

The adjusted R-squared value says that the treatment variable only explains 29% of the variance.  Nevertheless, we have plenty of statistical power supporting the treatment variable acting as a predictor for *weightGain* in this simple linear regression.

```{r}
SRHmodel <- lm(Train$SRHimprove ~ Train$trt)
summary(SRHmodel)
```
We see that the treatment variable explains about 4.5% of the total SRH improvement variance.  The average SRH improvement is 1/2 of a category.  We also have very strong statistical support.

### Multivariate Modeling of Weight Loss
We'll now build multivariate models using *sex*, *race*, and *treatment* to see how much better our model can be.  We'd also like to see if we can understand anything about those who are actually gaining weight as the result of the treatment.
```{r}
weightMVR <- lm(weightGain ~ sex * race * trt,data=Train)
summary(weightMVR)
```
We see that overall men lose about 2 pounds more than women, when their treatment/control status is not included.  We see that those in the treatment group lose 3 pounds compared to the control group, when no other attributes are included.  
We also see a specific Male:Treatment interaction term that is quite large.  Men who are treated lose an additional 10 pounds above and beyond the 2 pound loss of being men and the 3 pound loss of being treated.  
All 3 of these coefficients are statistically significant.  
The variables and all the interaction terms now explain about 65% of the weightGain variance.  
I'm going to follow this up with a plot, separating out men and women.
```{r}
ggplot(Train, aes(x=sex, y=weightGain), colour=trt) +
  geom_jitter(aes(colour=as.factor(trt)), size=0.5) + 
  ggtitle("Weight Gain Change for Women and Men") +
  scale_color_manual(name="Treatment",
                      breaks = c("0","1"),
                      values = palette,
                      labels = c("No", "Yes"))
```
This plot is actually quite remarkable, as it shows that virtually all weight gain that happens, happens to treated women.  It also shows that men respond much better in terms of weight loss--both the percentage of men that lose weight and the magnitude of the weight loss.    

### Refining Model
If we take sex as a known and build separate models for women and for men, I suspect that the simple linear regression model for men will be extremely predictive.

```{r}
Train_men <- Train %>% filter(sex == "MALE")
modelMen <- lm(weightGain ~ trt,data=Train_men)
summary(modelMen)
```
The treatment is associated with a 13.6 pound weight loss in men.

### Modeling Women
```{r}
Train_women <- Train %>% filter(sex == "FEMALE")
modelWomen <- lm(weightGain ~ trt, data=Train_women)
summary(modelWomen)
```
When the model looks only at women, there is an average weight loss of about 3 pounds, and it's quite statistically significant.  

I'm going to further tease apart the Women dataset to see if I can isolate those who are gaining weight.  
```{r}
modelWomenMVR <- lm(weightGain ~ trt * race * age, data=Train_women)
summary(modelWomenMVR)
```
We see that the coefficient for treated women who are black is +3.6 pounds, but it's not statistically significant.  Age seems to have virtually no impact.

It should be noted that each coefficient represents how much the weightGain variable changes per single unit change in the predictor, while holding all other predictors constant.  However, the predictors can be correlated with one another, and it may not be possible to only change the predictor in question while not changing any other predictors.

Next, I'd like to plot out the women by race
```{R}
ggplot(Train_women, aes(x=race, y=weightGain), colour=trt) +
  geom_jitter(aes(colour=as.factor(trt)), size=0.5) +
  ggtitle("Weight Gain for Women by Race") +
  scale_color_manual(name="Treatment",
                      breaks = c("0","1"),
                      values = palette,
                      labels = c("No", "Yes"))
```
Based on the plot, I believe that the magnitude of the treatment:Black-women coefficient may be due to a single outlier. (It wasn't statistically significant anyway.)  Overall, race doesn't help sub-divide the group of women who gain weight.

Plotting women's age:
```{r}
ggplot(Train_women, aes(x=age, y=weightGain), colour=trt) +
  geom_jitter(aes(colour=as.factor(trt)), size=0.5) +
  ggtitle("Weight Gain of Women Plotted vs. Age") +
  scale_color_manual(name="Treatment",
                      breaks = c("0","1"),
                      values = palette,
                      labels = c("No", "Yes"))
  
```
Again, consistent with the *modelWomenMVR* above, age is not an explanatory variable, with the possible exception of women under 25.


### SRH Multi-Variate Regression
```{r}
SRH_MVR <- lm(SRHimprove ~ sex * trt, data=Train)
summary(SRH_MVR)
```
Here, we're seeing that the treatment improves SRH overall, and there's an additional effect for treated men.  

Increasing age improves SRH improvement, but with a small magnitude.
Being hispanic or white also explains part of the improvement with larger magnitude but less statistical significance.

```{r}
ggplot(Train, aes(x=trt, y=SRHimprove), colour=sex) + 
  geom_jitter(aes(colour=sex), size=0.5) +
  theme_bw() + ggtitle("Self Reported Health Improvement") +
  scale_color_manual(name="Category",
                      breaks = c("MALE", "FEMALE"),
                      values = c("green", "blue"),
                      labels = c("Male", "Female"))
```
Once can clearly see the higher density of treated men compared to treated women with a positive SRH.

I'm going to now build separate SRH models for men and women.
```{r}
SRHmen <- lm(SRHimprove ~ trt, data=Train_men)
summary(SRHmen)
```
```{r}
SRHwomen <- lm(SRHimprove ~ trt, data=Train_women)
summary(SRHwomen)
```


# Conclusions:
1. The treatment is very effective for weight loss in men, with an average loss of over 13 pounds.  The result is statistically signficant.  
2. The treatment gives rise to an average weight loss of 3 pounds for women.  The result is statistically significant.  The treatment sometimes gives rise to weight gain of about XXX percent of women.
3. Neither race nor age is a significant predictor for weight loss.
4. The treatment gave rise to a half-unit improvement in self-reported health index in both men and women, but the effect improvement in men was almost twice as large as for women.
